---
title: "Introduction to hrfals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to hrfals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
```

The `hrfals` package implements Confound-Adjusted Alternating Least Squares
(CF-ALS) for estimating hemodynamic response functions (HRFs) and
condition amplitudes from fMRI time series. It integrates tightly with
the `fmridesign` ecosystem for constructing event models and HRF bases.

This short vignette walks through a minimal end-to-end example.

## Key ideas

- Confound-adjusted: projects out nuisance variation before HRF estimation.
- Alternating least squares: alternates between HRF coefficients and
  condition amplitude updates with regularization.
- Flexible bases: works with any HRF basis from `fmrihrf` (e.g., FIR,
  SPMG, FLOBS).

## Installation

Install from GitHub (requires `remotes`):

```r
# install.packages("remotes")
remotes::install_github("bbuchsbaum/hrfals")
```

The package also relies on `fmridesign` and `fmrihrf` which will be pulled
in automatically.

## Minimal workflow

1. Define a sampling frame and event model with `fmridesign`.
2. Choose an HRF basis from `fmrihrf`.
3. Fit with `hrfals()` (or `hrfals_sparse()` for L1-penalized betas).

```{r message=FALSE}
library(hrfals)
library(fmridesign)

# 1) Sampling frame and simple event model
TR <- 1
n_time <- 80
sf <- sampling_frame(blocklens = n_time, TR = TR)

events <- data.frame(
  onset = c(5, 20, 40, 60),
  condition = factor(c("A", "A", "B", "B")),
  block = 1
)

emod <- event_model(onset ~ hrf(condition), data = events,
                    block = ~ block, sampling_frame = sf)

# 2) Choose an HRF basis
hrf_basis <- fmrihrf::HRF_SPMG3

# 3) Simulate a small BOLD matrix (time x voxels)
Y <- matrix(rnorm(n_time * 5, sd = 0.2), n_time, 5)

# Fit CF-ALS
fit <- hrfals(Y, emod, hrf_basis,
              lam_beta = 10,   # beta ridge
              lam_h = 1,       # HRF ridge
              max_alt = 1)     # number of ALS alternations

fit
```

The returned object is of class `hrfals_fit` with fields including HRF
coefficients (`h_coeffs`), condition amplitudes (`beta_amps`), a
reconstructed HRF (`recon_hrf`), and goodness-of-fit (`gof`).

## Inspecting results

```{r}
str(list(
  h_dim = dim(fit$h_coeffs),
  beta_dim = dim(fit$beta_amps),
  gof_summary = summary(as.numeric(fit$gof))
))
```

If you want sparse condition amplitudes (useful when many conditions or
predictors are present), use `hrfals_sparse()` which applies an L1
penalty to `beta` by default:

```{r}
fit_sparse <- hrfals_sparse(Y, emod, hrf_basis, max_alt = 1)
str(dim(fit_sparse$beta_amps))
```

## Tips

- Start with small `max_alt` (e.g., 1–2) and moderate ridge penalties
  (`lam_beta`, `lam_h`), then tune.
- Use FIR bases for flexible shapes; use canonical bases for stability
  and interpretability.
- Provide confounds via `confound_obj` and a baseline model via
  `baseline_model` for real data to improve robustness.

## Learn more

- Build projected designs and inspect metadata: `vignette("design_and_confounds", package = "hrfals")`.
- Derive trial-wise beta series after fitting CF-ALS: `vignette("trialwise_beta_series", package = "hrfals")`.
- Tune sparse penalties for large predictor sets: `vignette("tuning_sparse_penalty", package = "hrfals")`.
- See `vignette("many_continuous_predictors", package = "hrfals")` for a
  sparse CF-ALS example with many predictors.
- Explore function docs: `?hrfals`, `?hrfals_sparse`,
  `?estimate_hrf_cfals`.
