---
title: "Estimating Shared HRFs with Many Continuous Predictors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Shared HRFs with Many Continuous Predictors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates the sparse CF-ALS implementation for scenarios
with a large number of continuous predictors.  The helper function
`hrfals_sparse()` provides sensible defaults for these use cases.

## Simulated example

The code below simulates a small data set with many predictors and then
fits the model using `hrfals_sparse()`.  The same workflow applies when
using real fMRI data and a design created with `fmridesign::event_model()`.

```{r eval=FALSE}
library(hrfals)
library(fmridesign)

n  <- 60   # time points
k  <- 100  # number of continuous predictors
h_len <- 5 # length of FIR basis

# simulate predictors
X_list <- lapply(seq_len(k), function(i) matrix(rnorm(n * h_len), n, h_len))

# sparse beta pattern
beta_true <- c(rep(1, 5), rep(0, k - 5))

# true HRF coefficients
h_true <- rnorm(h_len)

# generate BOLD signal
Y <- matrix(0, n, 1)
for (i in seq_len(k)) {
  Y <- Y + (X_list[[i]] %*% h_true) * beta_true[i]
}
Y <- Y + matrix(rnorm(n, sd = 0.1), n, 1)

# fit using sparse CF-ALS
fit <- cf_als_engine(
  X_list, Y,
  lambda_b = 0.1,
  lambda_h = 0.1,
  lambda_init = 0.1,
  R_mat_eff = diag(h_len),
  fullXtX_flag = FALSE,
  precompute_xty_flag = TRUE,
  Phi_recon_matrix = diag(h_len),
  h_ref_shape_canonical = rep(1, h_len),
  max_alt = 2,
  beta_penalty = list(l1 = 0.05, alpha = 1, warm_start = TRUE),
  design_control = list(standardize_predictors = TRUE,
                        cache_design_blocks = TRUE)
)

plot(fit)
```

The estimated HRF and sparse beta weights can be inspected from the
returned `fit` object.  When working with real fMRI data you would
construct `X_list` via `create_cfals_design()` which supports
splitting a single `Ident` term into multiple predictor blocks.
