---
title: "Tuning Sparse Betas in CF-ALS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning Sparse Betas in CF-ALS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
```

When many predictors compete for limited variance it is helpful to
encourage sparse condition amplitudes.  `hrfals()` supports elastic-net
penalties through the `beta_penalty` argument, and the helper
`tune_beta_l1_hrfals()` provides a lightweight grid search to pick a
reasonable L1 strength.  This vignette walks through a compact example.

## Simulate a sparse design

We construct an event model with ten conditions but only three of them
carry signal.  The remaining conditions should shrink toward zero when a
suitable penalty is used.

```{r}
library(hrfals)
library(fmridesign)

TR <- 1
n_time <- 100
sf <- sampling_frame(blocklens = n_time, TR = TR)

n_conditions <- 10
cond_names <- paste0("cond", seq_len(n_conditions))

set.seed(1)
pool <- seq(4, n_time - 10, by = 2)
onsets_all <- sort(sample(pool, size = n_conditions * 3, replace = FALSE))
events <- data.frame(
  onset = onsets_all,
  condition = factor(rep(cond_names, each = 3), levels = cond_names),
  block = 1
)

emod <- event_model(
  onset ~ hrf(condition),
  data = events,
  block = ~ block,
  sampling_frame = sf
)

hrf_basis <- fmrihrf::HRF_FIR  # default 12-column FIR basis

# ground-truth HRF basis coefficients and sparse beta vector
h_coeffs <- c(1.2, 0.8, 0.2, rep(0, fmrihrf::nbasis(hrf_basis) - 3))
beta_true <- c(rep(1, 3), rep(0, n_conditions - 3))

# obtain design blocks for simulation
cf_design <- create_fmri_design(emod, hrf_basis)
X_list <- cf_design$X_list

Y <- matrix(0, n_time, 20)
for (j in seq_along(X_list)) {
  signal_j <- drop(X_list[[j]] %*% h_coeffs) * beta_true[j]
  Y <- Y + signal_j
}
Y <- Y + matrix(rnorm(n_time * 20, sd = 0.25), n_time, 20)
```

## Run the penalty grid search

We try a small set of L1 penalty strengths while keeping the ridge
penalty fixed.  The helper splits voxels into train/test subsets and
returns the mean-squared error for each setting.

```{r}
tune_res <- tune_beta_l1_hrfals(
  fmri_data_obj_subset = Y,
  event_model = emod,
  hrf_basis = hrf_basis,
  l1_grid = c(1e-3, 5e-3, 1e-2, 5e-2),
  other_hrfals_args = list(lam_beta = 0.05, lam_h = 0.05),
  n_outer_iterations_cfals = 1
)

tune_res
attr(tune_res, "best_l1")
```

With a selected penalty it is straightforward to rerun `hrfals()` on the
full data using `beta_penalty = list(l1 = attr(tune_res, "best_l1"))`.
The grid search is intentionally simple—feel free to extend it with more
voxels or outer iterations when applying the package to real data.
